`phyr`: An R package for phylogenetic species-distribution modelling in ecological communities

### 2020-07-15 ms.Rmd
new intro

Ecological communities are collections of species that occur within the same geographical area. Which species occur within communities depends on the dispersal ability of species to enter the community, the environmental conditions that they find there, and the interactions that they have with other species in the community. These three processes -- dispersal, environmental tolerance, and species interactions -- depend on the traits that species possess and hence reflect evolutionary history and biogeographic processes [@warren2014mistaking; @gerhold2018deep]. For example, the larvae of an aquatic insect species might only occur in a lake if its adult stage has long-distance flight capabilities, if it can tolerate the low pH of the lake, and if it can avoid the predators that are common. Because traits play a central role in the composition of species that make up a community, community composition will likely reflect, at least in part, phylogenetic relationships among species. For example, two closely related insects might have similar dispersal capability, pH tolerance, and predator avoidance behavior, making them more likely to occur in the same lake. The recognition that phylogenetic relationships can increase our understanding of communities has led to a growing number of statistical methods for analyzing phylogenetic community composition [@ives2011generalized; @li2017can; @frishkoff2017phylogenetic; @losos1996; @webb2000; @webb2002phylogenies; @cavender2004; @helmus2007].

Just as the distributions of two species might reflect their proximity on a phylogenetic tree, the species occurring at two sites might reflect the sites’ geographical proximity. The most immediate possible cause of spatial correlations in species distributions is dispersal, if nearby sites are more likely to be colonized by a species. Spatial proximity may also be a surrogate for environmental variables that are unknown or unmeasured. For example, an insect species might occur in two nearby lakes because they both have low pH, yet pH has not been measured. Just as phylogenetic relationships among species can generate correlations between species in which sites they occupy, so too can spatial proximity generate correlations between sites in the species they contain [@cressie1991; @ives2006statistics].

How species respond to environmental factors, and how they respond to each other, depend on their traits. Therefore, correlations among functional traits can provide insights about the evolutionary history that has shaped species traits so they can occupy the same sites. For example, two insect species that occur in the same lake might share both long-range flight abilities and tolerance to low pH. Is the positive correlation between these two traits caused by correlated selective forces? A challenge to answering this question is that phylogenetic correlations between trait values might reflect species phylogenetic relatedness rather than shared selection: two species might have both long-range flight abilities and tolerance to low pH only because they are phylogenetically closely related. To distinguish between these two explanations -- convergence of suites of traits due to shared selective forces versus similarity due to phylogenetic relatedness -- it is necessary to account for phylogenies when performing correlation analyses between traits that could explain similarities in the distributions of species.

Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities [@ives2011generalized; @ovaskainen2011; @warton2015so]. The models can describe complex relationships in the data, such as how phylogenetically related species might respond similarly to the same environmental gradient, or how phylogenetically related species might exclude each other from the same communities. They also give a firm statistical basis to test these patterns, the ability to simulate data sets from the fitted model, and the ability to predict the composition of unsurveyed communities. These benefits of phylogenetic community composition models come with costs: building models can be intricate and fitting them computationally slow.

The R package `phyr` is designed to overcome many of these costs with a user-friendly interface, flexibility to build a rich collection of models, and good computational performance. Below, we first give a brief overview of the structure and syntax of two key functions `pglmm()` and `cor_phylo()`. `pglmm()` allows the formulation of a diverse set of phylogenetic generalized linear models (PGLMM) that can be used not only to analyze phylogenetic community composition but also comparative models for Gaussian and non-Gaussian data. `cor_phylo()` computes the Pearson correlations among species traits while simultaneously estimating the strength of phylogenetic signal within each trait. We then compare `pglmm()` and `cor_phylo()` to methods and programs that are currently available. Finally, we apply  `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.

>>>>

Ecological communities are collections of species that occur within the same geographical area. Which species occur within communities depends on the dispersal ability of species to enter the community, the environmental conditions that they find there, and the interactions that they have with other species in the community. For example, the larvae of an aquatic insect species might only occur in a particular lake if its adult stage has long-distance flight capabilities, if it can tolerate the low pH of the lake, and if it can avoid the predators that are common. These three processes -- dispersal, environmental tolerance, and species interactions -- depend on the traits that species possess and hence reflect evolutionary history and biogeographic processes [@warren2014mistaking; @gerhold2018deep]. Differences in community composition among locations depends on the variation in abiotic and biotic characteristics of the locations [@whittaker1956vegetation]. Continuing with our example of aquatic insects, an insect species might occur in two nearby lakes because they both have low pH, or because they have the same predator community. Nearby lakes might also have the same insect species if the adult insects readily disperse between lakes. Thus, understanding the composition of a community, and the variation in composition among communities, requires understanding both species traits and environmental variables that together govern what species can occur where, and whether they can get there.

The importance of species traits, environmental characteristics, and geographical location for community composition imply that statistical analyses of community composition should be capable of incorporating all of these types of factors. For any statistical analysis, a model specifies variables that might explain the presence or abundance of species (predictor or independent variables) and the statistical structure of the unexplained variation. Even though species traits may ultimately dictate whether a species can occur in a specific community, it is unlikely that all relevant traits are known and measured for every species. Consequently, the effects of these traits will be treated as "unexplained variance" in the model. Nonetheless, because phylogenetically related species often have similar trait values, the unexplained variance will likely have phylogenetic signal [@felsenstein1985phylogenies; @freckleton2002phylogenetic; @blomberg2003testing]. Therefore, statistical models of community composition should account for phylogenies [@li2017can; @frishkoff2017phylogenetic; @losos1996; @webb2000; @webb2002phylogenies; @cavender2004; @helmus2007]. Similarly, environmental factors differing among locations that are unknown or unmeasured might nonetheless be more similar among locations that are close to each other. This will generate spatial autocorrelation in the unexplained variance among communities [@cressie1991; @ives2006statistics]. Spatial autocorrelation can also be generated by the dispersal of individuals among nearby communities [@moran1953statistical; @royama2012analytical]. Thus, the analysis of community composition requires statistical models that can explicitly include species traits and environmental factors, and also include them implicitly in the form of phylogenetic relationships among species and geographical distances among locations in the unexplained variance terms of the model.

Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities [@ives2011generalized; @ovaskainen2011; @warton2015so]. The models can describe complex relationships in the data, such as how phylogenetically related species might respond similarly to the same environmental gradient, or how phylogenetically related species might exclude each other from the same communities. They also give a firm statistical basis to test these patterns, the ability to simulate data sets from the fitted model, and the ability to predict the composition of unsurveyed communities.

Analyzing the composition of communities in a phylogenetic context may generate additional statistical questions about the evolutionary processes that have generated the distribution of values of traits among species. For example, two insect species that occur in the same lake might share both long-range flight abilities and tolerance to low pH. Is the positive correlation between these two traits caused by correlated selective forces? A challenge to answering this type of question is that phylogenetic correlations between trait values might reflect species phylogenetic relatedness rather than shared selection: two species might have both long-range flight abilities and tolerance to low pH only because they are phylogenetically closely related. To distinguish between these two explanations -- convergence of suites of traits due to shared selective forces versus similarity due to phylogenetic relatedness -- it is necessary to account for phylogenies when performing correlation analyses between traits that could explain similarities in the distributions of species.

The R package `phyr` is designed to allow users to easily develop a rich collection of models for the analyses of ecological communities that include phylogenetic correlations among species and/or spatial correlation among locations. Below, we first give a brief overview of the structure and syntax of two key functions `pglmm()` and `cor_phylo()`. `pglmm()` allows the formulation of a diverse set of phylogenetic generalized linear models (PGLMM) that can be used not only to analyze phylogenetic community composition but also comparative models for Gaussian and non-Gaussian data. `cor_phylo()` computes the Pearson correlations among species traits while simultaneously estimating the strength of phylogenetic signal within each trait. We then compare `pglmm()` and `cor_phylo()` to methods and programs that are currently available. Finally, we apply  `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.
